# Module 2

## 2A. Data Management Part 1

We will practice plotting data using the iris dataset.

We will use the iris dataset for this section

```{r}
data(iris) # load data (already exists in base R)
```

```{r}
head(iris) # print first 6 lines of dataset
tail(iris) # print last 6 lines of dataset
```

```{r}
str(iris) # print 'structure' of dataset giving you info about each column
```

### Making and modifying variables

Here's how we make a new column that is a unique number:
```{r}
iris$Plant <- 1:length(iris$Species)
```

Here's how we make a new column that is total petal and sepal length:
```{r}
iris$PetSep.Length <- iris$Petal.Length+iris$Sepal.Length
```

Here's how to make a new column that log-transforms PetSep.Length:
iris$lnPS.Len <- log(iris$PetSep.Length)
```{r}
iris$lnPS.Len <- log(iris$PetSep.Length)
```

Here's how to make a new column for 'genus'. The only values you want is "Iris":
```{r}
iris$Genus <- 'Iris'
```

Here's how to combine two columns:
```{r}
iris$GenSpp <- paste(iris$Genus, iris$Species, sep="_")
```



Here's how to change Species 'versicolor' to 'versi' in the GenSpp column:
```{r}
iris$GenSpp <- gsub('versicolor', 'versi', iris$GenSpp )  ## looks for 'versicolor' and replaces it with 'versi' in the column iris$Species
```

sub() can be used for replacement but will only do 1 replacement and gsub() can also be used for replacement but with all matching instances.

You can use gsub() to add genus name to species column (alternative to making new column and then pasting together).
```{r}
iris$GenSpp1 <- gsub('.*^', 'Iris_', iris$Species)
```


### Variables with the tidyverse

```{r}
library(tidyverse)      # load package tidyverse (install if needed)
library(viridis)
data(iris)               # reload iris to clear changes from above
iris1 <- as_tibble(iris) # load iris and convert to tibble
```

```{r}
glimpse(iris1)     ## similar to str(), just glimpses data
```

mutate() will allow you create and modify variables:
```{r}
iris1 <- iris1 %>% mutate(Plant=1:length(Species), 
                          PetSep.Length=Petal.Length+Sepal.Length, 
                          lnPS.Len=log(PetSep.Length), 
                          Genus='Iris', 
                          GenSpp=gsub('.*^', 'Iris_', Species)) 
        ## note that I am overwriting iris1. Use with caution
```

summarize() calculates means, sd, min, max, etc. on a dataset
```{r}
iris1 %>% summarize(mean(Petal.Length))  ## mean of Petal.Length in dplyr
```

```{r}
mean(iris1$Petal.Length) ## mean of Petal.Length in base R
```

Here we summarize lnPS.Len by Species with both Tidyverse and base R:
```{r}
means_PetLen1 <- iris1 %>% group_by(Species) %>%
  summarize(Petal.Length=mean(Petal.Length)) ## tidy code
```

```{r}
means_PetLen2 <- aggregate(Petal.Length~Species, FUN="mean", data=iris1) ## base R
```

Here we summarize multiple variables by species use summarize_all()
```{r}
means1 <- iris1 %>% 
  select(Sepal.Length, Sepal.Width, Petal.Length,
         Petal.Width,lnPS.Len, Species) %>%  
  group_by(Species) %>% 
  summarize_all(list(mean=mean,sd=sd,n=length))
means1
```

### Reshape data for better usability 

Here is how we reshape data from wide to long:
```{r}
iris_long <- iris1 %>% group_by(Species) %>% 
  pivot_longer(cols=c(Sepal.Length, Sepal.Width, Petal.Length,
                      Petal.Width, lnPS.Len), 
               names_to = 'Trait', 
               values_to = 'value')
head(iris_long)
```

We can calculate the mean, sd, and n for each Species X trait combo and then calculate SE:
```{r}
means2 <- iris_long %>% 
  group_by(Species,Trait) %>% 
  summarize(mean=mean(value), 
            sd=sd(value), 
            n=length(value)) %>% 
  mutate(se=sd/sqrt(n)) %>%
  filter(Trait!='lnPS.Len')
head(means2)
```

Note that the previous code could all be done in one piped command:
```{r}
means2a <- iris1 %>% 
  group_by(Species) %>% 
  pivot_longer(cols=c(Sepal.Length, Sepal.Width, Petal.Length,
                      Petal.Width, lnPS.Len), names_to = 'Trait',
               values_to = 'value') %>% 
  group_by(Species,Trait) %>% 
  summarize(mean=mean(value), 
            sd=sd(value),
            n=length(value)) %>% 
  mutate(se=sd/sqrt(n)) %>%
  filter(Trait!='lnPS.Len')
means2a
```

We can make plot, below are two plots to start with. One is ineffective and one is more effective.
```{r}
ggplot(data=means2, aes(x=Species, y=mean, fill=Trait)) + 
  geom_point(size=5, position=position_dodge(width=0.25), pch=22) +
  labs(y="Floral part measurement (mm)") +
  geom_errorbar(aes(ymin=(mean-sd), ymax=(mean+sd)), width=.2,
                position=position_dodge(width=0.25), lwd=1.5) +
  scale_fill_viridis(discrete = T, 
                     labels=c("Petal Length","Petal Width",
                              "Sepal Length", "Sepal Width"),
                     option="magma") +
  theme(panel.border=element_rect(color="black",size=2, fill=NA)) +
  xlab("Species")
```

```{r}
ggplot(data=iris_long %>% 
         filter(Trait!='lnPS.Len'), aes(x=Species, 
                                        y=value, 
                                        fill=Species)) + 
  geom_boxplot() + 
  facet_wrap(~Trait, scales = 'free_y') +
  labs(y="Floral part measurement (mm)") +
  scale_fill_viridis(discrete = T, 
                     option = "plasma", 
                     direction = -1, begin=.2) +
  theme_bw()
```

## 2B. Data Management Part 2

We are still using the Iris data set as well as the tidyverse.
```{r}
library(tidyverse)
library(viridis)
iris1 <- as_tibble(iris) # load iris and convert to tibble
```


Here we make plot of sepal.length by sepal.width in wide format:
```{r}
ggplot(data=iris1, aes(x=Sepal.Width, y=Sepal.Length)) + 
  geom_point(color="#39568CFF") +   ## color outside of aes() changes color of all points (ie. not mapped to a column)
  facet_wrap(~Species) +
  geom_smooth(method='lm',color="#39568CFF", fill="#39568CFF")+
  theme_bw()
```

We can reshape data for for comparing traits in different panels:
```{r}
### reshape data from wide to long
iris_long <- iris1 %>% group_by(Species) %>% 
  pivot_longer(cols=c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width), names_to = 'Trait', values_to = 'value')
head(iris_long)
```

Now we make a plot comparing species with traits on different panels:
```{r}
ggplot(data=iris_long , aes(x=Species, y=value, fill=Species)) + 
  geom_boxplot() + 
  facet_wrap(~Trait, scales = 'free_y') +
  labs(y="Floral part measurement (mm)") +
  scale_fill_viridis(discrete = T, 
                     option = "plasma", 
                     direction = -1) +
  theme_bw()
```

## 2C. Data Exploration

For data exploration we need to load the following libraries:
```{r}
library(tidyverse) 
library(agridat)
library(corrplot) 
library(EnvStats) 
```

We will use the Iris dataset.
```{r}
data(iris) # load data (already exists in base R)
iris[8,3] <- 7 # plant data point for demo
head(iris) # print first 6 lines of dataset
tail(iris) # print last 6 lines of dataset
```

Here we use str() to print the 'structure' of dataset giving you info about each column or we can use glimpse().
```{r}
str(iris)
```
```{r}
glimpse(iris) # glimpse is similar to str() in tidyverse
```

### Distributions & summary statistics

We can view histograms of petal lengths with this:
```{r}
ggplot(iris, aes(x = Petal.Length)) + 
  geom_histogram(bins=12, color="white") +
  theme_bw(base_size = 16) +
  geom_vline(aes(xintercept = mean(Petal.Length)),
             color = "blue",
             size = 2) +
  geom_vline(aes(xintercept= median(Petal.Length)), 
             color = "orange", 
             size = 2)
```

We can also facet the histograms:
```{r}
ggplot(iris, aes(x = Petal.Length)) + 
  geom_histogram(bins=12, color="white") + 
  facet_wrap(~Species, scales="free") + 
  theme_bw(base_size = 16)

```

We can use the summary() to examine mean, median, and ranges. 
```{r}
summary(iris)
```

We can also get a table of means and medians with this tidyverse code:
```{r}
iris %>%
  pivot_longer(cols=c(1:4)) %>% 
  group_by(Species,name) %>% 
  summarize(mean=mean(value),median=median(value)) 
```

### Examining for outliers 

Boxplots can be used to examine distribution and look for outliers:
```{r}
ggplot(iris, aes(x=Species, y = Petal.Length)) + 
  geom_boxplot(fill="grey", width=.5) + 
  facet_wrap(~Species, scales="free") + 
  theme_bw(base_size = 16)
```

We can also use a dixon test for outliers or other tests like the grubbs test or Rosner test (for multiple outliers):
```{r}
library(outliers)
```

```{r}
## grubbs test for outliers, highest then lowest. Other functions EnvStats::rosnerTest() can test for multiple outliers
grubbs.test(iris$Petal.Length) ## full dataset
grubbs.test(iris$Petal.Length[iris$Species=='setosa']) ## just species setosa
grubbs.test(iris$Petal.Length[iris$Species=='setosa'], opposite=T) ## test lower outlier for species setosa
```

Here we can remove outliers and remake boxplots. Filtering with "|" (OR) will select all observations where one condition is met but not the other. 
```{r}
iris1 <- iris %>% filter(Petal.Length<4 | !Species=='setosa')
```

Ploting data:
```{r}
ggplot(iris1, aes(x=Species, y = Petal.Length)) + 
  geom_boxplot(fill="grey", width=.5) + 
  facet_wrap(~Species, scales="free") + 
  theme_bw(base_size = 16)
```

### Explore relationships

We can first use the GGally package. The ggpairs() code provides us with scatter plots that plot variables against one another in a pairwise fashion. We also see the distribution of the data and the correlation coefficients between a pair of variables

```{r}
library(GGally) ## install and load GGally package, if necessary
```

```{r}
ggpairs(iris1)   ## Make a big panel plot for exploration!!!
ggpairs(iris1, aes(color=Species, alpha=.75)) ## add color to seperate by species
```

Alternative to ggpairs() is the cor() which can be better for quickly scanning complex datasets:
```{r}
iris_cor <- cor(iris1 %>% select(-Species) %>% as.matrix()) ## first make correlation matrix
corrplot(iris_cor, method = "circle", type = "upper") ## plots strength of correlation as color-coded circles

```


